---
title: "Lab 09: SQL"
author: "Katie Zhao"
format: html
editor: visual
embed-resources: true
---

## **Intro**

In this lab, you will be working with a hospital database. There will be four tables that we will be working with: **doctor**, **visit**, **patient**, and **payment**. The main goal of the lab is to gain experience working with SQL databases. First, we will set up our SQL database:

```{r}
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")

# Copy data.frames to database
dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)
dbListTables(con)
```

TIP: You can use the following QUERY to see the structure of a table

```{sql connection=con}
PRAGMA table_info(doctor)
```

# **Exercise 1**

Edit the code below to retrieve the doctor ID, first name and last name for all doctors using the `doctor` table. Sort by last name and then by first name (note that the code chunk below is set up to run SQL code rather than R code).

```{sql connection=con}

SELECT doctor_id, first_name, last_name
FROM doctor
ORDER by last_name, first_name
```

# **Exercise 2**

Retrieve the doctor ID, first name, and last name for doctors whose last name equals ‘WILLIAMS’ or ‘DAVIS’.

```{sql connection=con}
SELECT doctor_id, first_name, last_name
FROM doctor
WHERE last_name IN ('WILLIAMS', 'DAVIS')

```

# **Exercise 3**

Write a query against the `visit` table that returns the IDs of the patients who visited a hospital starting on July 5, 2005 (use the visit_start_date column, and you can use the `date()` function to ignore the time component). Include a single row for each distinct patient ID.

```{sql connection=con}
SELECT DISTINCT visit_id
FROM visit
WHERE date(visit_start_date) = '2005-07-05'
```

## **Exercise 4**

**4.1) Construct a query that retrieves all columns of rows from the `payment` table where the amount is either 199, 799, or 999.**

```{sql connection=con}
SELECT *
FROM payment
WHERE amount IN (199, 799, 999)

```

**4.2) Construct a query that retrieves all columns of rows from the `payment` table where the amount is greater then 500.**

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 500

```

## **Exercise 4.3**

Construct a query that retrieves all columns of rows from the `payment` table where the amount is greater then 500 and less then 800.

```{sql connection=con}
SELECT *
FROM payment
WHERE amount > 500 AND amount < 800
```

# **Exercise 5**

Retrieve all the payment IDs and their amounts from the patients whose last name is ‘DAVIS’.

```{sql connection=con}
SELECT payment_id, amount
FROM patient AS a INNER JOIN payment AS b 
ON a.patient_id = b.patient_id
WHERE last_name = "DAVIS"

```

# **Exercise 6**

## **Exercise 6.1**

Use `COUNT(*)` to count the number of rows in `visit`.

```{sql connection=con}

SELECT COUNT(*) as RowTotal
FROM visit

```

## **Exercise 6.2**

Use `COUNT(*)` and `GROUP BY` to count the number of visits for each `patient_id`.

```{sql connection=con}

SELECT patient_id, COUNT(*) as PatientVisit
FROM visit
GROUP BY patient_id
```

## **Exercise 6.3**

Repeat the previous query and sort by the count in descending order.

```{sql connection=con}

SELECT patient_id, COUNT(*) as NumPatientVisit
FROM visit
GROUP BY patient_id
ORDER BY NumPatientVisit DESC

```

## **Exercise 6.4**

Repeat the previous query but use `HAVING` to only keep the groups with 40 or more.

```{sql connection=con}

SELECT patient_id, COUNT(*) as NumPatientVisit
FROM visit
GROUP BY patient_id
HAVING NumPatientVisit >= 40
ORDER BY NumPatientVisit DESC

```

# **Exercise 7**

Write a query that calculates a number of summary statistics for `amount` in the payment table using `MAX`, `MIN`, `AVG` and `SUM`

```{sql connection=con}

SELECT
  MAX(amount) AS max_amount, MIN(amount) AS min_amount,
  AVG(amount) AS avg_amount, SUM(amount) AS sum_amount
FROM payment

```

## **Exercise 7.1**

Modify the above query to do those calculations for each `patient_id`.

```{sql connection=con}

SELECT patient_id,
  MAX(amount) AS max_amount, MIN(amount) AS min_amount,
  AVG(amount) AS avg_amount, SUM(amount) AS sum_amount
FROM payment
GROUP BY patient_id

```

## **Exercise 7.2**

Modify the above query to only keep the `patient_id`s that have more then 5 payments.

```{sql connection=con}

SELECT patient_id,
  MAX(amount) AS max_amount, MIN(amount) AS min_amount,
  AVG(amount) AS avg_amount, SUM(amount) AS sum_amount
FROM payment
GROUP BY patient_id
HAVING COUNT(*) > 5

```

## **Exercise 8.1**

Create a SQL query to join the `doctor` and `visit` tables. Then, use SQL to compute the average visit length for each doctor. Hint: use the SQLite function `julianday` to convert a text date into a number.

```{sql connection=con}

SELECT first_name, last_name,
AVG(julianday(visit_end_date) - julianday(visit_start_date)) AS AvgVisitLength
FROM doctor AS a INNER JOIN visit AS b 
ON a.doctor_id = b.doctor_id
GROUP BY first_name, last_name

```

## **Exercise 8.2**

What are the names of the doctors with the 5 longest and 5 shortest average visits? Hint: use the `LIMIT` command.

```{sql connection=con}
SELECT first_name, last_name,
AVG(julianday(visit_end_date) - julianday(visit_start_date)) AS AvgVisitLength
FROM doctor AS a INNER JOIN visit AS b 
ON a.doctor_id = b.doctor_id
GROUP BY first_name, last_name
ORDER BY AvgVisitLength DESC
LIMIT 5
```

```{sql connection=con}
SELECT first_name, last_name,
AVG(julianday(visit_end_date) - julianday(visit_start_date)) AS AvgVisitLength
FROM doctor AS a INNER JOIN visit AS b 
ON a.doctor_id = b.doctor_id
GROUP BY first_name, last_name
ORDER BY AvgVisitLength
LIMIT 5
```

# **Cleanup**

Run the following chunk to disconnect from the connection.

```{r}
dbDisconnect(con)
```
